Concentration of Maxima and Fundamental Limits in High-Dimensional Testing and Inference

### 2021-02-01 6.GWAS/6.optimal-design.tex
move proofs of GWAS results: 1. move proofs of phase transition results to appendix. 2. move proofs of results on signal sizes in text

Corollary \ref{cor:optimal-design} is proved in Section \ref{subsec:proof-signal-size-odds-ratio}.

>>>>

\begin{proof}
	Using the parametrization in \eqref{eq:reparametrize-2-by-2-table-1}, we solve for $\delta$ in \eqref{eq:odds-ratio-delta} to obtain
	\begin{align}
		\delta &= \frac{\phi_1 fR}{fR+1-f} - \left(\frac{\phi_1 fR}{fR+1-f} + f(1-\phi_1)\right)\phi_1 \nonumber \\
		&= \frac{f(1-f)\phi_1(1-\phi_1)(R-1)}{fR+1-f}. \label{eq:reparametrize-2-by-2-table-2}
	\end{align}
	Substituting \eqref{eq:reparametrize-2-by-2-table-2} into the expression \eqref{eq:signal-size-chisq-delta}, after some simplification, yields
	\begin{equation} \label{eq:reparametrize-2-by-2-table-3}
	w^2 = \frac{f(1-f)\phi_1(1-\phi_1)(R-1)^2}{\left[\phi_1 R + (1-\phi_1)D\right]\left[\phi_1 + (1-\phi_1)D\right]},
	\end{equation}
	where $D = fR+1-f > 0$.
	Therefore, he derivative of \eqref{eq:reparametrize-2-by-2-table-3} with respect to $\phi_1$ is
	\begin{equation} \label{eq:signal-size-first-derivative}
	\frac{\mathrm{d}w^2}{\mathrm{d}\phi_1} = 
	\frac{f(1-f)(R-1)^2}{\left[\phi_1 R+(1-\phi_1)D\right]^2 \left[\phi_1+(1-\phi_1)D\right]^2} \left[(D^2-R)\phi_1^2 - 2D^2\phi_1 + D^2\right].
	\end{equation}
	Further, we obtain the second derivative with respect to $\phi_1$,
	\begin{equation} \label{eq:signal-size-second-derivative}
	\frac{\mathrm{d}^2w^2}{\mathrm{d}\phi_1^2} = 
	h(R,f) \left[(\phi_1-1)D^2 - \phi_1R\right],
	\end{equation}
	where $h$ is some function of $(R,f)$ taking on strictly positive values.

	Since $\left[(\phi_1-1)D^2 - \phi_1R\right]<0$, the second derivative \eqref{eq:signal-size-second-derivative} must be strictly negative on $[0,1]$.
	This implies that the first derivative \eqref{eq:signal-size-first-derivative} is strictly decreasing on $[0,1]$. 
	Since the first derivative \eqref{eq:signal-size-first-derivative} is strictly positive at $\phi_1=0$, and strictly negative at $\phi_1=1$, it must have a unique zero between 0 and 1, and hence, the solution to $(D^2-R)\phi_1^2 - 2D^2\phi_1 + D^2 = 0$ in the interval of $[0,1]$ must be the maximizer of \eqref{eq:reparametrize-2-by-2-table-3} --- when $D^2-R>0$, the smaller of the two roots maximizes \eqref{eq:reparametrize-2-by-2-table-3}, and when $D^2-R<0$, it is the larger of the two.
	They share the same expression ${D}/{(D+\sqrt{R})}$, which coincides with \eqref{eq:optimal-design}.
	Finally, when $D^2=R$, the only root $\phi_1^*=1/2$, which also coincides with \eqref{eq:optimal-design}, is the maximizer of \eqref{eq:reparametrize-2-by-2-table-3}.
\end{proof}

### 2021-02-01 6.GWAS/6.proofs-GWAS.tex
move proofs of GWAS results: 1. move proofs of phase transition results to appendix. 2. move proofs of results on signal sizes in text

\subsection{Auxiliary facts of chi-square distributions}
\label{sec:chi-square-distributions}
\input{6.GWAS/6.chi-square-distributions.tex}

\subsection{Proof of Theorem \ref{thm:chi-squared-exact-boundary}}
\label{subsec:proof-chi-squared-exact-boundary}

@@

\subsection{Proof of Theorem \ref{thm:chi-squared-approx-boundary}}
\label{subsec:proof-chi-squared-approx-boundary}

@@

\subsection{Proof of Theorems \ref{thm:chi-squared-exact-approx-boundary} and \ref{thm:chi-squared-approx-exact-boundary}}
\label{subsec:proof-chi-squared-mix-boundaries}

@@

\subsection{Proof of Proposition \ref{prop:signal-size-odds-ratio} and Corollary \ref{cor:signal-limits-OR}}
\label{subsec:proof-signal-size-odds-ratio}

\begin{proof}[Proof of Proposition \ref{prop:signal-size-odds-ratio} and Corollary \ref{cor:signal-limits-OR}]
We parametrize the 2-by-2 multinomial distribution with the parameter $\delta$, 
\begin{equation} \label{eq:reparametrize-2-by-2-table-1}
    \mu_{11} = \phi_1\theta_1+\delta,\quad 
    \mu_{12} = \phi_1\theta_2-\delta,\quad 
    \mu_{21} = \phi_2\theta_1-\delta,\quad 
    \mu_{22} = \phi_2\theta_2+\delta.
\end{equation}
By relabelling of categories, we may assume $0<\theta_1,\phi_1\le1/2$ without loss of generality.
Note that $\delta$ must lie within the range $[\delta_\mathrm{min}, \delta_\mathrm{max}]$, where
$$
\delta_\mathrm{min} := \max\{-\phi_1\theta_1, -\phi_2\theta_2, \phi_1\theta_2-1, \phi_2\theta_1-1\} 
= -\phi_1\theta_1,
$$
and
$$
\delta_\mathrm{max} := \min\{1-\phi_1\theta_1, 1-\phi_2\theta_2, \phi_1\theta_2, \phi_2\theta_1\}
= \min\{\phi_1\theta_2, \phi_2\theta_1\},
$$
in order for $\mu_{ij}\ge0$ for all $i,j\in \{1,2\}$.
Under this parametrization, Relation \eqref{eq:odds-ratio} then becomes
\begin{equation} \label{eq:odds-ratio-delta}
    \text{R} = \frac{\mu_{11}\mu_{22}}{\mu_{12}\mu_{21}}
    = \frac{\phi_1\theta_1\phi_2\theta_2 + \delta(\phi_1\theta_1+\phi_2\theta_2)+\delta^2}{\phi_1\theta_1\phi_2\theta_2 - \delta(\phi_1\theta_2+\phi_2\theta_1)+\delta^2},
\end{equation}
which is one-to-one and increasing in $\delta$ on $(\delta_\mathrm{min}, \delta_\mathrm{max})$.
Equation \eqref{eq:signal-size-chisq} becomes
\begin{equation} \label{eq:signal-size-chisq-delta}
w^2 = \sum_{i=1}^2 \sum_{j=1}^2 \frac{(\mu_{ij} - \phi_i\theta_j)^2}{\phi_i\theta_j}
= \delta^2\sum_i\sum_j \frac{1}{\phi_i\theta_j}
= \frac{\delta^2}{\phi_1\theta_1\phi_2\theta_2},
\end{equation}
Solving for $\delta$ in \eqref{eq:odds-ratio-delta}, and plugging into the expression for signal size \eqref{eq:signal-size-chisq-delta} yields Relation \eqref{eq:signal-size-odds-ratio}.
Corollary \ref{cor:signal-limits-OR} follows from the fact that $w^2(\delta)$ is decreasing on $[\delta_\mathrm{min},0)$, increasing on $(0,\delta_\mathrm{max}]$, with limits
$$
\lim_{d\to \delta_\mathrm{min}} w^2(\delta) = \frac{\phi_1\theta_1}{\phi_2\theta_2},
\quad
\text{and}
\quad
\lim_{d\to \delta_\mathrm{max}} w^2(\delta) = \min\left\{\frac{\phi_1\theta_2}{\phi_2\theta_1}, \frac{\phi_2\theta_1}{\phi_1\theta_2}\right\}.
$$
The other three cases ($1/2\le\theta_1,\phi_1\le1$; $0<\theta_1\le1/2\le\phi_1\le1$; and $0\le\phi_1\le1/2\le\theta_1\le1$) may be obtained similarly, or by appealing to the symmetry of the problem.
\end{proof}

\begin{proof}[Proof of Corollary \ref{cor:optimal-design}]
Using the parametrization in \eqref{eq:reparametrize-2-by-2-table-1} and in Corollary \ref{cor:signal-size-odds-ratio-conditional-frequency}, we solve for $\delta$ in \eqref{eq:odds-ratio-delta} to obtain
\begin{align}
    \delta &= \frac{\phi_1 fR}{fR+1-f} - \left(\frac{\phi_1 fR}{fR+1-f} + f(1-\phi_1)\right)\phi_1 \nonumber \\
    &= \frac{f(1-f)\phi_1(1-\phi_1)(R-1)}{fR+1-f}. \label{eq:reparametrize-2-by-2-table-2}
\end{align}
Substituting \eqref{eq:reparametrize-2-by-2-table-2} into the expression \eqref{eq:signal-size-chisq-delta}, after some simplification, yields
\begin{equation} \label{eq:reparametrize-2-by-2-table-3}
    w^2 = \frac{f(1-f)\phi_1(1-\phi_1)(R-1)^2}{\left[\phi_1 R + (1-\phi_1)D\right]\left[\phi_1 + (1-\phi_1)D\right]},
\end{equation}
where $D = fR+1-f > 0$.
Therefore, he derivative of \eqref{eq:reparametrize-2-by-2-table-3} with respect to $\phi_1$ is
\begin{equation} \label{eq:signal-size-first-derivative}
    \frac{\mathrm{d}w^2}{\mathrm{d}\phi_1} = 
    \frac{f(1-f)(R-1)^2}{\left[\phi_1 R+(1-\phi_1)D\right]^2 \left[\phi_1+(1-\phi_1)D\right]^2} \left[(D^2-R)\phi_1^2 - 2D^2\phi_1 + D^2\right].
\end{equation}
Further, we obtain the second derivative with respect to $\phi_1$,
\begin{equation} \label{eq:signal-size-second-derivative}
    \frac{\mathrm{d}^2w^2}{\mathrm{d}\phi_1^2} = 
    h(R,f) \left[(\phi_1-1)D^2 - \phi_1R\right],
\end{equation}
where $h$ is some function of $(R,f)$ taking on strictly positive values.

Since $\left[(\phi_1-1)D^2 - \phi_1R\right]<0$, the second derivative \eqref{eq:signal-size-second-derivative} must be strictly negative on $[0,1]$.
This implies that the first derivative \eqref{eq:signal-size-first-derivative} is strictly decreasing on $[0,1]$. 
Since the first derivative \eqref{eq:signal-size-first-derivative} is strictly positive at $\phi_1=0$, and strictly negative at $\phi_1=1$, it must have a unique zero between 0 and 1, and hence, the solution to $(D^2-R)\phi_1^2 - 2D^2\phi_1 + D^2 = 0$ in the interval of $[0,1]$ must be the maximizer of \eqref{eq:reparametrize-2-by-2-table-3} --- when $D^2-R>0$, the smaller of the two roots maximizes \eqref{eq:reparametrize-2-by-2-table-3}, and when $D^2-R<0$, it is the larger of the two.
They share the same expression ${D}/{(D+\sqrt{R})}$, which coincides with \eqref{eq:optimal-design}.
Finally, when $D^2=R$, the only root $\phi_1^*=1/2$, which also coincides with \eqref{eq:optimal-design}, is the maximizer of \eqref{eq:reparametrize-2-by-2-table-3}.
\end{proof}

>>>>

\section{Auxiliary facts of chi-square distributions}
\label{sec:chi-square-distributions}
\input{6.GWAS/6.chi-square-distributions.tex}

\section{Proof of Theorem \ref{thm:chi-squared-exact-boundary}}
\label{subsec:proof-chi-squared-exact-boundary}

@@

\section{Proof of Theorem \ref{thm:chi-squared-approx-boundary}}
\label{subsec:proof-chi-squared-approx-boundary}

@@

\section{Proof of Theorems \ref{thm:chi-squared-exact-approx-boundary} and \ref{thm:chi-squared-approx-exact-boundary}}
\label{subsec:proof-chi-squared-mix-boundaries}

@@
